A comparative analysis of penalized linear mixed models in structured genetic data

### 2020-09-08 main.tex
3.5 edits

This result also allows us to mathematically justify the statement from section \ref{sec:background}, that SNP-phenotype associations will be unbiased in the presence of environmental heterogeneity which does not correspond to population structure. We can quantify the proportion of the environmental effect which is in the column space of $\boldsymbol{P}_{\bX}$ \anna{and thus biases the $\bbeta$ estimates} using the ratio \anna{$||\bX \boldsymbol{\tau}||_2^2 / ||\bZ \bgamma||_2^2 \in [0, 1]$}. In the case where $\bZ \bgamma$ is entirely independent of population structure, $\bZ \bgamma \in \mathcal{N}(\boldsymbol{P}_{\bX})$ and $\bX \boldsymbol{\tau} = \boldsymbol{0}$ so that $\mathbb{E}(\bbetaHat^*) = \bbeta$. Thus the SNP-phenotype associations will be unbiased, but the residual variance will be increased. Although we focus our attention on accurately estimating fixed SNP effects, we note that this result could lead to substantial bias in narrow-sense heritability estimates.

In practical applications $\bZ \bgamma$ is unlikely to be entirely orthogonal to population structure. However, the degree to which it reflects population structure that can be inferred from $\bX$, as quantified by \anna{$||\bX \boldsymbol{\tau}||_2^2 / ||\bZ \bgamma||_2^2$}, will impact the extent to which environmental heterogeneity \anna{has the potential to bias} SNP-phenotype associations. \anna{We will use this metric in Section \ref{sec:results}. Given the high-dimensional nature of our data, we will approximate $(\bX^\T \bX)^{-1}$ using a low-rank approximation. We use this approach as opposed to that of a generalized inverse in order to ensure a unique solution.}

@@

\subsubsection{Design matrix generation}
We generated these pseudophenotypes using three types of high-dimensional ($p > n$) $\bX$ matrices: (1) simulated SNP data from independent subpopulations, (2) simulated SNP data from admixed subpopulations, and (3) real genome-wide SNP data from known caucasian ($n = 417$), African American ($n = 134$), and Hispanic ($n = 37$) subpopulations \anna{cite Vanderwas data}. $p = 1000$, $n = \{200, 400, 600, 800 \}$ for the simulated data types, and 50 causal SNPs were used throughout.

@@

We begin by comparing the ordinary Lasso, PC-Lasso, and LMM-Lasso as the level of environmental confounding is varied. Estimation accuracy for lasso, PC-Lasso, and LMM-Lasso was evaluated based on mean squared error (MSE) and broken down into bias and variance components. Figure \ref{fig:mse} presents these results for all three data types in a 1:1 signal to noise (SNR) setting, where $\eta = 0.5$. The top row of Figure \ref{fig:mse} shows that when $\xi = 0$ (i.e., no environmental confounding), lasso and LMM-Lasso perform comparably. PC-Lasso results in a slight inflation in MSE compared to the other methods due to the fact that 10 additional predictors have been added to the model, but none of them actually explain any variability in the outcome. The bottom row of Figure \ref{fig:mse} shows the MSE when $\xi = 0.8$. In this case, both PC-Lasso and LMM-Lasso clearly outperform the naive lasso. This improvement is more substantial in the independent subpopulations simulated data than the admixed simulated data. \anna{quantify this for all 3 data settings using $||\bX \boldsymbol{\tau}||_2 / ||\bZ \bgamma||_2$: Admixture ratio = 0.46412, Independent ratio = 0.98615, Vanderwas ratio = 0.64325} This is due to the fact that the discrete, nonoverlapping genetic subpopulations more directly correspond to the distinct fixed environmental effects than do the more continuous and blended subpopulations in the admixture data. In other words, a greater amount of the environmental effect is in the column space of the projection matrix of $\bX$ for the independent subpopulations data than in the admixed data. As such, there is a greater amount of confounding and potential estimation bias in the independent subpopulations which can be corrected by the PC-Lasso and LMM-Lasso methods.

>>>>

This result also allows us to mathematically justify the statement from section \ref{sec:background}, that SNP-phenotype associations will be unbiased in the presence of environmental heterogeneity which does not correspond to population structure. We can quantify the proportion of the environmental effect which is in the column space of $\boldsymbol{P}_{\bX}$ using the ratio 
\begin{equation}
    \label{eqn:ratio}
    \anna{||\bX \boldsymbol{\tau}||_2^2 / ||\bZ \bgamma||_2^2 \in [0, 1]}.
\end{equation} 
In the case where $\bZ \bgamma$ is entirely independent of population structure, $\bZ \bgamma \in \mathcal{N}(\boldsymbol{P}_{\bX})$ and $\bX \boldsymbol{\tau} = \boldsymbol{0}$ so that $\mathbb{E}(\bbetaHat^*) = \bbeta$. Thus the SNP-phenotype associations will be unbiased, but the residual variance will be increased. Although we focus our attention on accurately estimating fixed SNP effects, we note that this result could lead to substantial bias in narrow-sense heritability estimates.

In practical applications $\bZ \bgamma$ is unlikely to be entirely orthogonal to population structure. However, the degree to which it reflects population structure \anna{and has the potential to bias SNP-phenotype associations can be inferred from $\bX$ and quantified by \anna{\eqref{eqn:ratio}}. We will use this metric in our analysis of our simulation results in Section \ref{sec:results}. Given the high-dimensional nature of our data, we will approximate $(\bX^\T \bX)^{-1}$ using a low-rank approximation. We use this approach as opposed to that of a generalized inverse in order to ensure a unique solution.}

@@

\subsubsection{Design matrix generation}
We generated these pseudophenotypes using three types of high-dimensional ($p > n$) $\bX$ matrices: (1) simulated SNP data from independent subpopulations, (2) simulated SNP data from admixed subpopulations, and (3) real genome-wide SNP data from known caucasian ($n = 417$), African American ($n = 134$), and Hispanic ($n = 37$) subpopulations \anna{cite Vanderwas data PATRICK TO DO}. $p = 1000$, $n = \{200, 400, 600, 800 \}$ for the simulated data types, and 50 causal SNPs were used throughout.

@@

We begin by comparing the ordinary Lasso, PC-Lasso, and LMM-Lasso as the level of environmental confounding is varied. Estimation accuracy for lasso, PC-Lasso, and LMM-Lasso was evaluated based on mean squared error (MSE) and broken down into bias and variance components. Figure \ref{fig:mse} presents these results for all three data types in a 1:1 signal to noise (SNR) setting, where $\eta = 0.5$. The top row of Figure \ref{fig:mse} shows that when $\xi = 0$ (i.e., no environmental confounding), lasso and LMM-Lasso perform comparably. PC-Lasso results in a slight inflation in MSE compared to the other methods due to the fact that 10 additional predictors have been added to the model, but none of them actually explain any variability in the outcome. The bottom row of Figure \ref{fig:mse} shows the MSE when $\xi = 0.8$. In this case, both PC-Lasso and LMM-Lasso clearly outperform the naive lasso.

\anna{Using the metric $||\bX \boldsymbol{\tau}||_2 / ||\bZ \bgamma||_2$}

\anna{quantify this for all 3 data settings using $||\bX \boldsymbol{\tau}||_2 / ||\bZ \bgamma||_2$: Admixture ratio = 0.46412, Independent ratio = 0.98615, Vanderwas ratio = 0.64325}

This improvement is more substantial in the semi-empirical and independent subpopulations data than the admixed data. This is due to the fact that the discrete, nonoverlapping genetic subpopulations more directly correspond to the distinct fixed environmental effects than do the more continuous and blended subpopulations in the admixture data. In other words, a greater proportion of the environmental effect is in the column space of the projection matrix of $\bX$ for the semi-empirical and independent subpopulations data than in the admixed data.

As such, there is a greater amount of confounding and potential estimation bias in the independent subpopulations which can be corrected by the PC-Lasso and LMM-Lasso methods.
